An indexed construction of semi-simplicial and semi-cubical types

### 2022-05-23 paper/paper.tex
paper: adjust margins, lscape two tables, polish abstract
paper: minor polish

\usepackage{multirow, minted, tabularx, booktabs, setspace, makecell, caption, xstring, pdflscape}
\usepackage[a4paper, inner=25mm, outer=25mm, top=25mm, bottom=25mm]{geometry}

@@

% More magic: we extend xkeyval with a \appendmask macro to pass to \setkeys
\makeatletter
\edef\bonak@mask{}

\def\bonak@appendmask[#1]{
  \XKV@addtolist@n\bonak@mask{\g@addto@macro{\XKV@prefix\XKV@tfam}{@}#1}
}
\def\appendmask{\XKV@testopta{\XKV@testoptc\bonak@appendmask}}
\makeatother

% Now we patch XKV@setkeys
\makeatletter
\let\XKV@setkeys@alt\XKV@setkeys
\long\def\XKV@setkeys[#1]#2{
  \edef\bonak@pfx{\g@addto@macro{\XKV@prefix\XKV@tfam}{@}}
  \edef\bonak@tempa\@empty
  \edef\bonak@keys{}
  \XKV@for@n{\bonak@mask}\bonak@tempa{
    \IfBeginWith{\bonak@tempa}{\bonak@pfx}{
      \XKV@addtolist@n\bonak@keys{\StrBehind{\bonak@tempa}{\bonak@pfx}}}{}
  }
  % Finally, call the real function
  \XKV@setkeys@alt[\bonak@keys]{#2}
}
\makeatother

@@

% A block in the eqntable environment
\newcommand{\eqnline}[4]{$#1$ & $#2$ & $#3$ & $#4$ \\}
\newcommand{\mc}[1]{\multicolumn{4}{c}{\textit{#1}} \\\\}

@@

\begin{document}
\thispagestyle{empty}
\maketitle
\begin{abstract}
  A model of homotopy type theory, that formalizes augmented semi-simplicial sets that naturally generalizes to semi-cubical sets as well, in dependent type theory. The accompanying artifact is an axiomless~\footnote{Technically speaking, we use some axioms related to functional extensionality, but the requirement arises only from our generalization to $\nu$-types; if $\nu$ is finite (say, unit type corresponding to augmented semi-simplicial types), this requirement vanishes.} Coq formalization that uses sophisticated rewriting techniques and pushes the boundaries of proof assistant technology.
\end{abstract}
\tableofcontents
\newpage

@@

The outline of the paper is as follows. First, we take the opportunity to familize the reader with simplicial sets, as it appears in mathematical literature, along with a short section on semi-cubical sets. We then proceed to explain our construction, with the help of diagrams. Before spelling out the details of our formalization in type-theoretic language, we have a section on the unicity of identity proofs (UIP), since the use of \mintinline{coq}{SProp} has greatly eased our Coq formalization.

@@

\section{Our construction}
The Grothendieck construction provides a correspondence between a fibered representation, and an indexed representation.

$$
  \begin{tikzcd}
    X_0 : \U{} & X_1 : \U{} \arrow[l, "\delta^\epsilon_0" description, shift left=2] \arrow[l, "\delta^\omega_0" description, shift right=2] & X_2 : \U{} \arrow[l, "\delta^\epsilon_1" description, shift left=6] \arrow[l, "\delta^\epsilon_0" description, shift left=2] \arrow[l, "\delta^\omega_0" description, shift right=2] \arrow[l, "\delta^\omega_1" description, shift right=6] & \ldots
  \end{tikzcd}
$$

\begin{align*}
  X_0 & : \U{}                                                                                                      \\
  X_1 & : X_0 \times X_0 \rightarrow \U{}                                                                           \\
  X_2 & : \forall a b c d, X_1 : ab \rightarrow X_1 : bc \rightarrow X_1 : cd \rightarrow X_1 : da \rightarrow \U{} \\
  \ldots
\end{align*}

@@

One way of formalising set theory in type theory is via Aczel sets [ref]. It can be written down as an inductive as follows:

\begin{listing}[H]
  \begin{minted}[mathescape]{coq}
    Inductive SET : Type :=
      node : ~$\forall$~A : Type, (A -> SET) -> SET
      ~$=_\textrm{SET}$~ : ...
  \end{minted}
\end{listing}

\mintinline{coq}{(A -> SET) -> SET} can be drawn as:

\begin{equation*}
  \begin{tikzcd}
    \{0\} \arrow[dr] & \\
    & \{\} \\
  \end{tikzcd}
  \begin{tikzcd}
    & \{0, 1\} \arrow[dr]\arrow[dl] & \\
    \{0\} & & \{1\} \\
  \end{tikzcd}
  \begin{tikzcd}
    & & \{0, 1, 2\} \arrow[dr]\arrow[dl] & \\
    & \{0, 1\} \arrow[dl]\arrow[dr] & & \{2\} \\
    \{0\} & & \{1\} & \\
  \end{tikzcd}
\end{equation*}

In general, \mintinline{coq}{SET} can have an $n$-ary branching structure, splitting out the various subsets until we get to the leaves.

Then, proof irrelevance can be stated as:

\begin{align*}
  \forall x y, \forall p q : x =_{\textrm{SET}} y, p = q
\end{align*}

@@

The definition is dispatched over tables \ref{tab:core}, \ref{tab:frames}, \ref{tab:faces} and \ref{tab:coh}. It describes the structure of the underlying higher-dimensional relations on which fillers are built. Note the presence of a coherence condition $\cohf[l]$ to ensuring that both sides of the equality in $\restrl[l]$ and $\restrc[l]$ are in the same type. The proof of $\cohf[l]$ itself requires an higher-dimensional coherence condition which we obtain by working in \mintinline{coq}{HSet}. Note that if the proofs of the same equality were not equated, there would be a need for arbitrary many higher-dimensional coherences (see e.g.~\cite{Herbelin15} for a discussion on the de facto need for recursive higher-dimensional coherence conditions in formulating higher-dimensional structures in type theory). Note also that for a given $n$, the coherence conditions evaluate to a reflexivity proof, so that the construction evaluates to an effective sequence of types of iterated relations not mentioning $\restrf[l]$ nor $\cohf[l]$ anymore.

When reflexivities are excluded, we call the structure thus defined \emph{bare truncated sets}: \emph{bare} because it can be seen as defining truncated $n$-sets corresponding to simplical sets with only faces as part of the structure.

@@

% From the third table onwards, make the {D, E} arguments of prims implicit
\appendmask[prim]{frame}[D]
\appendmask[prim]{layer}[D]
\appendmask[prim]{filler}[D, E]

@@

\newpage
\begin{thebibliography}{10}
  \bibitem[Bez]{Bezem14}
  Bezem, M., Coquand, T., \& Huber, S. (2014, July). A model of type theory in cubical sets. \textit{In 19th International Conference on Types for Proofs and Programs (TYPES 2013)} (Vol. 26, pp. 107-128). Wadern, Germany: Schloss Dagstuhl–Leibniz Zentrum fuer Informatik.

  \bibitem[CubAgda]{Vezzosi21}
  Vezzosi, A., Mörtberg, A., \& Abel, A. (2021). Cubical Agda: a dependently typed programming language with univalence and higher inductive types. \textit{Journal of Functional Programming, 31}.

  \bibitem[CohCoq]{Cohen16}
  Cohen, C., Coquand, T., Huber, S., \& Mörtberg, A. (2016). Cubical type theory: a constructive interpretation of the univalence axiom. arXiv preprint arXiv:1611.02108.

@@

  \bibitem[Fri]{Friedman08}
  Friedman, G. (2008). An elementary illustrated introduction to simplicial sets. \textit{arXiv preprint arXiv:0809.4221}.

  \bibitem[Rie]{Riehl11}
  Riehl, E. (2011). A leisurely introduction to simplicial sets. \textit{Unpublished expository article available online from the author's web page}.

  \bibitem[CubSet]{Antolini00}
  Antolini, R. (2000). Cubical structures, homotopy theory. \textit{Annali di Matematica pura ed applicata, 178}(1), 317-324.

  \bibitem[Her]{Herbelin15}
  Herbelin, H. (2015). A dependently-typed construction of semi-simplicial types. \textit{Mathematical Structures in Computer Science, 25}(5), 1116-1131.

  \bibitem[CoqInCoq]{Barras97}
  Barras, B., \& Werner, B. (1997). Coq in coq. \textit{Available on the WWW.}
\end{thebibliography}

>>>>

\usepackage{multirow, minted, tabularx, booktabs, setspace, makecell, caption, xkeyval, xstring, xpatch, pdflscape}
\usepackage[a4paper, inner=25mm, outer=25mm, top=25mm, bottom=25mm]{geometry}

@@

% More magic: we patch xkeyval's \setkeys to use a mask to ignore keys
\makeatletter
% Our global mask
\def\bonak@mask{}
\def\bonak@na{}

% First we define \appendmask
\long\def\bonak@appendmask[#1]{
  \XKV@for@eo\XKV@fams\XKV@tfam{
    \def\bonak@hdr{\XKV@prefix\XKV@tfam @}
    \XKV@for@o{#1}\XKV@tempa{
      \XKV@addtolist@x\bonak@mask{\bonak@hdr\XKV@tempa}
    }
  }
}
\def\appendmask{\XKV@testopta{\XKV@testoptc\bonak@appendmask}}

% Now we patch \XKV@setkeys
\xpretocmd{\XKV@setkeys}{
  \XKV@for@eo\XKV@fams\XKV@tfam{
    \def\bonak@hdr{\XKV@prefix\XKV@tfam @}
    \XKV@for@eo{\bonak@mask}\XKV@tempa{
      \IfBeginWith{\XKV@tempa}{\bonak@hdr}{
        \def\XKV@tempb{\StrBehind{\XKV@tempa}{\bonak@hdr}}
        \XKV@addtolist@o\bonak@na{\XKV@tempb}}{}
    }
  }
}{\tracingpatches}{\tracingpatches}
\xpatchcmd{\XKV@setkeys}{#1}{\bonak@na}{\tracingpatches}{\tracingpatches}
\makeatother

@@

% A block in the eqntable environment
\newcommand{\eqnline}[4]{$#1$ & $#2$ & $#3$ & $#4$ \\}
\newcommand{\mc}[1]{\multicolumn{4}{c}{\emph{#1}} \\\\}

@@

\begin{document}
\thispagestyle{empty}
\maketitle
\begin{abstract}
  A model of homotopy type theory, formalizing augmented semi-simplicial sets, that naturally generalizes to semi-cubical sets as well, in dependent type theory. The accompanying artifact is an axiom-less~\footnote{Technically speaking, we use some axioms related to functional extensionality, but the requirement arises only from our generalization to $\nu$-types; if $\nu$ is finite (say, unit type corresponding to augmented semi-simplicial types), this requirement vanishes.} Coq formalization that uses sophisticated rewriting techniques and pushes the boundaries of proof assistant technology.
\end{abstract}
\tableofcontents
\newpage

@@

The outline of the paper is as follows. First, we take the opportunity to familize the reader with simplicial sets, as it appears in mathematical literature, along with a short section on semi-cubical sets. We then proceed to explain our construction, with the help of diagrams. Before spelling out the details of our formalization in type-theoretic language, we have a section on reasoning in sets, in which we spell out \texttt{HSet}.

@@

\section{Our construction}
The fibered representation:

\begin{equation*}
  \begin{tikzcd}
    X_0 : \U{} & X_1 : \U{} \arrow[l, "\delta^\epsilon_0" description, shift left=2] \arrow[l, "\delta^\omega_0" description, shift right=2] & X_2 : \U{} \arrow[l, "\delta^\epsilon_1" description, shift left=6] \arrow[l, "\delta^\epsilon_0" description, shift left=2] \arrow[l, "\delta^\omega_0" description, shift right=2] \arrow[l, "\delta^\omega_1" description, shift right=6] & \ldots
  \end{tikzcd}
\end{equation*}

The indexed representation:
\begin{align*}
  X_0 & : \U{}                                                                                                      \\
  X_1 & : X_0 \times X_0 \rightarrow \U{}                                                                           \\
  X_2 & : \forall a b c d, X_1 : ab \rightarrow X_1 : bc \rightarrow X_1 : cd \rightarrow X_1 : da \rightarrow \U{} \\
  \ldots
\end{align*}

@@

@@

The definition is dispatched over tables \ref{tab:core}, \ref{tab:frames}, \ref{tab:faces} and \ref{tab:coh}. It describes the structure of the underlying higher-dimensional relations on which fillers are built. Note the presence of a coherence condition $\cohf$ to ensuring that both sides of the equality in $\restrl$ and $\restrc$ are in the same type. The proof of $\cohf$ itself requires an higher-dimensional coherence condition which we obtain by working in HSet. Note that if the proofs of the same equality were not equated, there would be a need for arbitrary many higher-dimensional coherences (see e.g.~\cite{Herbelin15} for a discussion on the de facto need for recursive higher-dimensional coherence conditions in formulating higher-dimensional structures in type theory). Note also that for a given $n$, the coherence conditions evaluate to a reflexivity proof, so that the construction evaluates to an effective sequence of types of iterated relations not mentioning $\restrf$ nor $\cohf$ anymore.

We call the structure thus defined \emph{bare truncated sets}: \emph{bare} because it can be seen as defining truncated $\nu$-sets corresponding to simplical sets with only faces as part of the structure.

@@

% From the third table onwards, make the {D, E} arguments of prims implicit
\appendmask[prim]{frame}[D]
\appendmask[prim]{layer}[D]
\appendmask[prim]{filler}[D,E]

@@

\newpage
\begin{thebibliography}{10}
  \bibitem[Bez]{Bezem14}
  Bezem, M., Coquand, T., \& Huber, S. (2014, July). A model of type theory in cubical sets. \emph{In 19th International Conference on Types for Proofs and Programs (TYPES 2013)} (Vol. 26, pp. 107-128). Wadern, Germany: Schloss Dagstuhl–Leibniz Zentrum fuer Informatik.

  \bibitem[CubAgda]{Vezzosi21}
  Vezzosi, A., Mörtberg, A., \& Abel, A. (2021). Cubical Agda: a dependently typed programming language with univalence and higher inductive types. \emph{Journal of Functional Programming, 31}.

  \bibitem[CohCoq]{Cohen16}
  Cohen, C., Coquand, T., Huber, S., \& Mörtberg, A. (2016). Cubical type theory: a constructive interpretation of the univalence axiom. \emph{arXiv preprint arXiv:1611.02108}.

@@

  \bibitem[Fri]{Friedman08}
  Friedman, G. (2008). An elementary illustrated introduction to simplicial sets. \emph{arXiv preprint arXiv:0809.4221}.

  \bibitem[Rie]{Riehl11}
  Riehl, E. (2011). A leisurely introduction to simplicial sets. \emph{Unpublished expository article available online from the author's web page}.

  \bibitem[CubSet]{Antolini00}
  Antolini, R. (2000). Cubical structures, homotopy theory. \emph{Annali di Matematica pura ed applicata, 178}(1), 317-324.

  \bibitem[Her]{Herbelin15}
  Herbelin, H. (2015). A dependently-typed construction of semi-simplicial types. \emph{Mathematical Structures in Computer Science, 25}(5), 1116-1131.

  \bibitem[CoqInCoq]{Barras97}
  Barras, B., \& Werner, B. (1997). Coq in coq. \emph{Available on the WWW.}
\end{thebibliography}
